Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-04 aistats/paper.tex
Clean up gaussian variance section

\paragraph{Conjugate Prior}
One conjugate prior for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
We can re-write this prior as 
\begin{align}
    p(\natp) \propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
\begin{multline}
    -\log p(\cD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
\end{multline}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
\end{align}
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

@@

\subsection{Gaussian Variance}
The trailing example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
\end{align}

We can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.

The logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
In particular,
\begin{align}
	p(\natp) 
	&\propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \\
    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$. After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.

Both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}

In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{bregmandef.pdf}
	\includegraphics[width=.4\textwidth]{phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

\begin{figure}[ht]
	\centering
\includegraphics[width=.48\textwidth]{asymptote.pdf}
\includegraphics[width=.48\textwidth]{fewsamples.pdf}
	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$. Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.  
	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
	\textbf{Right:} the first few samples significantly deviate from this behavior. In fact, for $n=1$ and $n=2$, the expected value is infinite, but we have a closed form solution and a simple upper-bound for $n>2$.
	}
	\label{fig:curves}
\end{figure}

@@

This upper bound is asymptotically tight.
We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0 +n}{a+ (n-2)_+}
		\label{eq:theta_expectation_bound}
	\end{equation}
	where $(x)_+ = \max(0,x)$.
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP.

\begin{theorem}[MAP Bound]
 Let us  define
 \begin{align}
	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}

This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.

\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
	\caption{
	Convergence rate of MAP~\eqref{eq:MAP_rate} 
	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
	We report these curves for various values of $n$, and $n_0=5$. 
	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$). 
	This observation hints towards the optimality of MLE for large $n$. 
	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
	}
\end{figure}

>>>>

\paragraph{Bregman Divergences}
A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
It is a non-symmetric measure of distance between parameters.
It is equal to the divergence of its convex conjugate with switched arguments
\begin{align}
	\bregman (\natp ; \natp_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
Then the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
\begin{multline}
    -\log p(\cD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst
    \label{eq:joint_likelihood}
\end{multline}
meaning the posterior given $\cD=(X_1,\dots, X_n)$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
\end{align}
verifying the first oder optimality conditions \RLP{$\natp$ unconstrained, living in an open set with barriers}
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
\end{align}
These rules are known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

@@

\subsection{Gaussian Variance}
A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$. 
The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other $\natp = -\inv{2 \mu}$.
Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
\begin{align}
	\logpart (\natp) &= - \half \log(-\natp)  + \half \log(\pi) \\
	\conj(\mu) &= \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* )
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z))
\end{align}
illustrated in Figure~\ref{fig:phi}. 
Below we report upper bounds on the expected value of this divergence for the MLE and the MAP.

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
	\label{fig:phi}
\end{figure}

\begin{figure}[ht]
	\centering
\includegraphics[width=.4\textwidth]{fewsamples.pdf}
	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$. Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval. 
		While the expected value is infinite for $n=1$ or $n=2$, it quickly matches the upper bound, and the $1/2n$ asymptote.
	}
	\label{fig:curves}
\end{figure}

@@

This upper bound is asymptotically tight.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
We get a similar bound for the multivariate generalization :
the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$ which implies
\begin{align}
	\label{eq:log_bound} 
	\phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{theorem}[MAP Bound]
 For $n\in \naturalnumbers$, let us  define
 \begin{align}
	b_n = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}

This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.
\RLP{Add numerical experiments with upper bound for gaussian variance MAP.}
